Computational biology in the 21st century:\\Scaling with compressive algorithms

### 2015-09-22 main.tex
Bonnie's edits

\keywords{computational biology, database search, metagenomics, chemogenomics}

@@

Computational biologists answer biological and biomedical 
questions by using computation in support of---or in place of---laboratory 
procedures, in hopes of obtaining more accurate answers at a greatly reduced 
cost.
The past two decades have seen unprecedented technological progress with regard
to generating biological data; next-generation sequencing, mass spectrometry,
microarrays, nuclear magnetic resonance spectroscopy, and other high-throughput 
approaches have led to an explosion of data.
However, this explosion of data is a mixed blessing.
On the one hand, the scale and scope of data should allow new insights into
genetic and infectious diseases, cancer, basic biology, and even human migration
patterns.
On the other hand, researchers are generating datasets so massive that it has 
become difficult to analyze them to discover patterns that give clues to the 
underlying biological processes.

Certainly, computers are getting faster and more economical; the amount of 
processing available per dollar of compute hardware is more or less doubling 
every year or two; a similar claim can be made about storage capacity 
(Figure~\ref{fig:growth}).
In 2002, when the first human genome was sequenced, the growth in computing 
power was still matching the growth rate of genomic data.
However, the sequencing technology used for the Human Genome Project--Sanger
sequencing--was supplanted around 2004, with the advent of what is now known as 
next-generation sequencing.
The material costs to sequence a genome have plummeted in the past
decade, to the point where a whole human genome can be sequenced for less than
one thousand U.S. Dollars.
As a result, the amount of genomic data available to 
researchers is increasing by a factor of ten every year.

@@

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
Currently, many biological `omics' applications require us to store, access, 
and analyze large libraries of data.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have collaborated to bring the GATK 
(Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
Amazon Web Services are also commonly used for computational biology 
research and enterprise (e.g., DNAnexus)~\cite{schatz2010cloud}.
However, while cloud computing frees researchers from maintaining their own
data centers and provides cost-saving benefits when computing resources are
not needed continuously, it is no panacea.
First and foremost, the computer systems that make up those cloud datacenters
are themselves bound by improvements in semiconductor technology and Moore's
law. Thus, cloud computing does not truly address the problem posed by the
faster-than-Moore's law exponential growth in omics data.
Moreover, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.

@@

In the central dogma of molecular biology, DNA is transcribed into RNA, which
is translated by the ribosome into polypeptide chains, sequences of amino 
acids which singly or in complexes are known as proteins.
Proteins fold into complex, low-energy structures, which function as
cellular machines; the DNA sequence determines the amino acid sequence,
which in turn determines the folded structure of a protein.
This structure ultimately determines a protein's function within the cell.
Certain kinds of RNA also function as cellular 
machines.
Methods have been developed to gather biological data from every level of this process,
resulting in a massive influx of data on sequence, abundance, structure, function, and interaction of DNA, RNA, and proteins.
Much of these data are amenable to standard Big Data analysis methods; however, in this Review we focus on examples of biological data which exhibit additional exploitable structure for creating scalable algorithms.

Sequence data, either 
nucleotide sequences (using a four-letter alphabet representing the four
DNA or RNA bases) or protein sequences (using a twenty-letter alphabet 
representing the twenty standard amino acids)
are obtained in several ways.
For protein and RNA sequence data, mass spectrometry---which
can determine protein sequence and interactions---and RNA-seq---which can determine RNA sequence and abundance, 
allowing scientists to also infer the expression of the gene to which it might translate---play central roles.
However, with the advent of next-generation sequencing (NGS) technologies, the greatest volume of sequence data available are that of DNA.
To better understand the structure of NGS sequence data, we will expand on NGS methodologies.

At the dawn of the genomic era, Sanger sequencing was the most widely-used 
method for reading a genome.
Though slow and labor intensive, Sanger sequencing has the advantage of being
well-tested and is still used to validate NGS results.
More recently, however, Sanger sequencing has largely been supplanted by NGS approaches,
beginning with Illumina's `sequencing by synthesis,' which allows vastly 
greater throughput due to massive parallelism, low cost, and simple sample
preparation.
Illumina sequencing and other NGS approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing do not read a single DNA
molecule end-to-end as one could read through a bound book.
Instead, in so-called `shotgun sequencing' DNA molecules are chopped into many small fragments;
from these fragments we generate 
\emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
These reads must be put together in the correct order 
to piece together an entire genome.
Current reads typically range from 50 to 200 bases long, though longer reads are
available with some technologies (e.g., PacBio).
Because no sequencing technology is completely infallible, sequencing machines also
provide a \emph{quality score} (or measure of the confidence in the DNA base
called) associated with each position.
Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
characters that encode the quality of the base call.
A sequencing run will produce many overlapping reads.

@@

When examining more complex phenomena than single proteins or compounds, we
often look to synthesize things together into a systems-level understanding of biology.
To that end, we frequently use
networks to represent biological data, such as the genetic
and physical interactions among proteins, as well as those in metabolic 
pathways~\cite{berger2013computational}.
While standard network science tools have been employed in these analyses---
e.g. several approaches make use of diffusion or random walks to explore the topology of networks \cite{cao2013going,cho2015diffusion}---
they are often paired with more specific biological data,
as seen in IsoRank~\cite{singh2008global} and IsoRankN's~\cite{liao2009isorankn} use of conserved biological function in addition to random walks for global multiple network alignment.
Other tools solve other biological problems, such as Mongoose~\cite{chindelevitch2014exact}, which 
analyzes metabolic networks.
However, given its breadth, biological network science is beyond the scope of this Review, so we will not discuss it further.

@@

Given DNA or RNA reads from NGS technologies, the first task is to 
\emph{assemble} those fragments of sequence into contiguous sequences.
The assembly problem is analogous to the problem of 
reconstructing a book with all its pages torn out.
\emph{de novo} assembly is beyond the scope of this Review, but is
possible because the sequence is covered by many
overlapping reads~\cite{berger2013computational}; for this task, the \emph{de Bruijn graph}
data structure is commonly used~\cite{bruijn1946combinatorial}.
Often, however, a \emph{reference genome} (or in the case of RNA, 
\emph{transcriptome}) is available for the organism being sequenced; the
establishment of a human reference genome was indeed the purpose of the Human
Genome Project.

When a reference sequence is available, NGS reads can be \emph{mapped} onto
this reference (Figure~\ref{fig:ngs-pipeline}).
Continuing the book analogy, it's much easier to reconstruct a book with all its pages torn out when one has another (perhaps imperfect) copy of that book to match pages to.
Mapping allows the differences between the newly-sequenced genome and the 
reference to be analyzed; these differences, or variants, may include single-nucleotide polymorphisms
(SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
insertions or deletions, or larger-scale changes in the genome.
Determining the differences between an individual genome and a reference is
known as \emph{variant calling}.
While reference-based read mapping is a fundamentally simpler problem than
\emph{de novo} assembly, it is still computationally complex, as gigabytes or
terabytes of reads must each be mapped onto the reference genome, which can 
range from millions (for bacteria) to billions (for mammals) of base pairs.
As an example, the ICGC-TCGA Pan Cancer Analysis of Whole Genomes 
(PCAWG)~\cite{weinstein2013cancer} brings together more than 500 top cancer 
researchers from about 80 institutions in a coordinated manner with the goal of 
mapping the entire mutational landscape of 37 common cancer types. 
Currently each sample requires seven hours to download even on an institutional 
connection. 
Importantly, researchers do not trust the provided mapping, and thus they redo mappings. 
The time spent on mapping is about 50\% of the overall time spent on the sequence analysis pipeline. 
As read mapping is typically the most costly step in NGS analysis pipelines (e.g., GATK), any improvement to existing mappers will immediately accelerate sequence analysis studies on large read datasets.

@@

Given a sequenced genome, the next natural questions ask what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{berger2013computational} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function is
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known. 
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.
The homology search problem was previously believed to be solved;
Basic Local Alignment Search Tool (BLAST)~\cite{berger2013computational} had been
the standard tool for performing homology (similarity) search on databases of nucleotide and
protein sequences.
BLAST takes a `seed-and-extend' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running-time scales linearly with the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

@@

Techniques for reference-based read mapping typically rely on algorithmic 
approaches such as the Burrows-Wheeler transform (BWT), 
which provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure is a compressed substring index, 
based on the BWT, which provides efficient storage as well as fast search~\cite{berger2013computational}.
BWA (Burrows-Wheeler Aligner)uses the BWT, while the 
Bowtie2 mapper further relies on the FM-index for 
efficient mapping of NGS reads~\cite{berger2013computational} .
The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
coupled with dynamic programming in a compressed representation of the 
reference genome, in order to prune the search space 
when mapping reads to a reference genome.
Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2010mrsfast} use an `approximate seed' approach to indexing
the space of possible matches, likewise pruning the search space; however, the
bulk of its runtime is spent on the extend phase.
State-of-the-art mapper mrsFAST-Ultra achieves improvements in efficiency based on machine architecture rather than leveraging redundancy in the data itself with near-perfect sensitivity, but
only for the case where there are no insertions and deletions (indels) \cite{hach2014mrsfast}.
Even with these approaches, read mapping remains a significant bottlenecks in
genomic research~\cite{berger2013computational}.

Compressing reads for storage is necessary should researchers wish to apply more advanced mapping tools or other analysis in the future.
Taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common short
substring) together into `buckets', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This approach allows a general-purpose compressor to achieve tight compression.
SCALCE~\cite{berger2013computational} also relies on a `boosting' scheme, 
reordering
reads in such a way that a general-purpose compressor achieves improved
compression.

Recent advances in metagenomic search tools have relied on two improvements over
BLASTX: indexing and alphabet reduction.
RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
collision-free hash table.
The alphabet reduction, as it is reversible, can be thought of as a form of
lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
alphabet, with offsets stored to recover the original sequence in the full
alphabet.
The hash-table provides an efficient index of the database to be searched.
DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
`shaped seeds'---essentially, $k$-mers with wildcards at specific positions---instead of simple $k$-mer seeds to index the database.
DIAMOND demonstrates search performance three to four orders of magnitude faster
than BLASTX, but it is still linear in the size of the database being searched.

@@

\begin{table}
\label{fractal}
\caption{Metric entropy ratio (ratio of clusters to entries in database) and
fractal dimension at typical search radii for four data sets. Metric entropy ratio gives an estimate of the acceleration of coarse search with respect to na\"ive search, and as long as fractal dimension is low, coarse search should dominate total search time.
NCBI's non-redundant `NR' protein and `NT' nucleotide sequence databases are from June, 2015. Protein Data Bank (PDB) is from July, 2015. PubChem is from October, 2013.}
\tabcolsep=0.11cm
\begin{tabular}{H{3.0cm}p{2.2cm}p{1.8cm}}
\hline
Data set & Metric entropy ratio & Fractal dimension \\        
\hline
Nucleotide~sequences (NCBI NT) & 7:1 & 1.5\\
\hline
Protein~sequences (NCBI NR) & 5:1 & 1.6\\
\hline
Protein~structure (PDB) & 10:1 & 2.5\\
\hline
Chemical~structure (PubChem) & 11:1 & 0.2\\
\hline
\end{tabular}
\end{table}

@@

We are entering the age of compressive algorithms, 
which makes use of this completely different paradigm for the structure of biological data.
Seeking to take advantage of the redundancy inherent in genomic sequence data, 
Loh, Baym and Berger \cite{loh2012compressive}
introduced \emph{compressive genomics}, an approach that relies on compressing
data in such a way that the desired computation (such as BLAST search) can be
performed in the compressed representation.
Daniels et al. \cite{daniels2013compressive} further demonstrated that
compressive acceleration takes advantage of increasing redundancy in the
data over time.

Compressive genomics is based on the concept of \emph{compressive acceleration},
which relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
Coarse search is performed only on the coarse, or representative, subsequences
that represent unique data.
Any representative sequence within some threshold of the query is then expanded 
into all similar sequences it represents; the fine search is over this
(typically small) subset of the original database.
This approach provides orders-of-magnitude run-time improvements to BLAST 
nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
search; these run-time improvements increase as databases grow.

The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST (compressively-accelerated BLAST)~\cite{loh2012compressive} and 
caBLASTP~\cite{daniels2013compressive}, accelerates existing tools (in this 
case, read
mappers including BWA or Bowtie2) by allowing them to operate in a compressed
space, and relying on a coarse and a fine phase.
In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{berger2013computational}, conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as `seeds' for longer sequence matches.
CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each 
one to its neighbors within a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.
As an aside, though not an example of accelerated similarity search, mid-size $l$-mers
can also be used for storage compression of NGS quality scores associated with base calls.
Unlike many other approaches to compressing quality scores in the literature
~\cite{bonfield2013compression,ochoa2013qualcomp},
Quartz~\cite{yu2015quality} takes advantage of the fact that mid-size $l$-mers can in many cases almost uniquely identify locations in the genome,
bounding the likelihood that a quality score is informative and allowing for lossy compression of uninformative scores.
Because Quartz's lossy compression injects information from the distribution of $l$-mers in the target genome, it demonstrates not only improved compression over competing
approaches, but slightly improves the accuracy of downstream variant-calling.

@@

The compressive omics approach can, in some cases, come at the cost of accuracy.
However, these cases are well-defined.
Compressive omics never results in false positives (with respect to the na\"ive 
search technique being accelerated), because the fine search phase applies the
same comparison to the candidates as the na\"ive approach.
Furthermore, when the distance function used for comparisons is a metric---more specifically, when it obeys the triangle inequality---false negatives will also
never occur.
However, in practice, non-metric distance functions are used, such as E-values
in BLAST or cosine distance in esFragBag, and thus false negatives can occur.
In practice, these error rates are low, and recall better than 90\% has been
demonstrated~\cite{loh2012compressive,daniels2013compressive,yu2015entropy}.

@@

The approach of compressive acceleration~\cite{loh2012compressive}, and its 
demonstrated ability to scale with the metric entropy of the 
data~\cite{yu2015entropy}, while providing orthogonal benefits to many other
useful indexing techniques, is an important tool for coping with the deluge of
data.
The extension of this compressive acceleration approach to 
metagenomics~\cite{yu2015entropy}, NGS read 
mapping and chemogenomics~\cite{yorukoglu2015compressive} suggests its 
flexibility.

>>>>

\keywords{computational biology, database search, metagenomics, \\chemogenomics}

@@

Computational biologists answer biological and biomedical 
questions by using computation in support of---or in place of---laboratory 
procedures, in hopes of obtaining more accurate answers at a greatly reduced 
cost.
The past two decades have seen unprecedented technological progress with regard
to generating biological data; next-generation sequencing, mass spectrometry,
microarrays, cryo-electron microscopy, and other high-throughput 
approaches have led to an explosion of data.
However, this explosion of data is a mixed blessing.
On the one hand, the scale and scope of data should allow new insights into
genetic and infectious diseases, cancer, basic biology, and even human migration
patterns.
On the other hand, researchers are generating datasets so massive that it has 
become difficult to analyze them to discover patterns that give clues to the 
underlying biological processes.

Certainly, computers are getting faster and more economical; the amount of 
processing available per dollar of compute hardware is more or less doubling 
every year or two; a similar claim can be made about storage capacity 
(Figure~\ref{fig:growth}).
In 2002, when the first human genome was sequenced, the growth in computing 
power was still matching the growth rate of genomic data.
However, the sequencing technology used for the Human Genome Project---Sanger
sequencing---was supplanted around 2004, with the advent of what is now known 
as next-generation sequencing.
The material costs to sequence a genome have plummeted in the past
decade, to the point where a whole human genome can be sequenced for less than
one thousand U.S. Dollars.
As a result, the amount of genomic data available to 
researchers is increasing by a factor of ten every year.

@@

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
Currently, many biological `omics' applications require us to store, access, 
and analyze large libraries of data.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have collaborated to bring the GATK 
(Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
Amazon Web Services are also commonly used for computational biology 
research and enterprise (e.g., DNAnexus)~\cite{schatz2010cloud}.
However, while cloud computing frees researchers from maintaining their own
data centers and provides cost-saving benefits when computing resources are
not needed continuously, it is no panacea.
First and foremost, the computer systems that make up those cloud datacenters
are themselves bound by improvements in semiconductor technology and Moore's
law. Thus, cloud computing does not truly address the problem posed by the
faster-than-Moore's law exponential growth in omics data.
Moreover, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained;
accessing cloud resources for analytics may not be possible.

@@

In the central dogma of molecular biology, DNA is transcribed into RNA, which
is translated by the ribosome into polypeptide chains, sequences of amino 
acids which singly or in complexes are known as proteins.
Proteins fold into sophisticated, low-energy structures, which function as
cellular machines; the DNA sequence determines the amino acid sequence,
which in turn determines the folded structure of a protein.
This structure ultimately determines a protein's function within the cell.
Certain kinds of RNA also function as cellular 
machines.
Methods have been developed to gather biological data from every level of this process,
resulting in a massive influx of data on sequence, abundance, structure, function, and interaction of DNA, RNA, and proteins.
Much of these data are amenable to standard Big Data analysis methods; however, in this Review we focus on examples of biological data which exhibit additional exploitable structure for creating scalable algorithms.

Sequence data, either 
nucleotide sequences (using a four-letter alphabet representing the four
DNA or RNA bases) or protein sequences (using a twenty-letter alphabet 
representing the twenty standard amino acids)
are obtained in several ways.
For both protein and RNA sequence data, mass spectrometry---which
can determine protein sequence and interactions---and RNA-seq---which can determine RNA sequence and abundance, 
allowing scientists to also infer the expression of the gene to which it might translate---play central roles.
However, with the advent of next-generation sequencing (NGS) technologies, the greatest volume of sequence data available are that of DNA.
To better understand the structure of NGS sequence data, we will expand on NGS methodologies.

At the dawn of the genomic era, Sanger sequencing was the most widely-used 
method for reading a genome.
More recently, however, NGS approaches,
beginning with Illumina's `sequencing by synthesis,' have enabled vastly 
greater throughput due to massive parallelism, low cost, and simple sample
preparation.
Illumina sequencing and other NGS approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing do not read a single DNA
molecule end-to-end as one could read through a bound book.
Instead, in so-called `shotgun sequencing' DNA molecules are chopped into many small fragments;
from these fragments we generate 
\emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
These reads must be put together in the correct order 
to piece together an entire genome.
Current reads typically range from 50 to 200 bases long, though longer reads are
available with some technologies (e.g., PacBio).
Because no sequencing technology is completely infallible, sequencing machines also
provide a \emph{quality score} (or measure of the confidence in the DNA base
called) associated with each position.
Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
characters that encode the quality of the base call.
A sequencing run will produce many overlapping reads.

@@

When examining more complex phenomena than single proteins or compounds, we
often look to synthesize things together into a systems-level understanding of biology.
To that end, we frequently use
networks to represent biological data, such as the genetic
and physical interactions among proteins, as well as those in metabolic 
pathways~\cite{berger2013computational}.
While standard network science tools have been employed in these analyses---
e.g. several approaches make use of diffusion or random walks to explore the topology of networks \cite{cao2013going,cho2015diffusion}---
they are often paired with more specific biological data,
as seen in IsoRank~\cite{singh2008global} and IsoRankN's~\cite{liao2009isorankn} use of conserved biological function in addition to random walks for global multiple network alignment.
Other tools solve other biological problems, such as MONGOOSE~\cite{chindelevitch2014exact}, which 
analyzes metabolic networks.
However, given its breadth, biological network science is beyond the scope of this Review, so we will not discuss it further.

@@

Given DNA or RNA reads from NGS technologies, the first task is to 
\emph{assemble} those fragments of sequence into contiguous sequences.
The assembly problem is analogous to the problem of 
reconstructing a book with all its pages torn out.
\emph{De novo} assembly is beyond the scope of this Review, but is
possible because the sequence is covered by many
overlapping reads~\cite{berger2013computational}; for this task, the \emph{de Bruijn graph}
data structure is commonly used~\cite{bruijn1946combinatorial}.
Often, however, a \emph{reference genome} (or in the case of RNA, 
\emph{transcriptome}) is available for the organism being sequenced; the
establishment of a human reference genome was indeed the purpose of the Human
Genome Project.

When a reference sequence is available, NGS reads can be \emph{mapped} onto
this reference (Figure~\ref{fig:ngs-pipeline}).
Continuing the book analogy, it is much easier to reconstruct a book with all its pages torn out when one has another (perhaps imperfect) copy of that book to match pages to.
Mapping allows the differences between the newly-sequenced genome and the 
reference to be analyzed; these differences, or variants, may include single-nucleotide polymorphisms
(SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
insertions or deletions, or larger-scale changes in the genome.
Determining the differences between an individual genome and a reference is
known as \emph{variant calling}.
While reference-based read mapping is a fundamentally simpler problem than
\emph{de novo} assembly, it is still computationally complex, as gigabytes or
terabytes of reads must each be mapped onto the reference genome, which can 
range from millions (for bacteria) to billions (for mammals) of base pairs.
As an example, the ICGC-TCGA Pan Cancer Analysis of Whole Genomes 
(PCAWG)~\cite{weinstein2013cancer} brings together more than 500 top cancer 
researchers from about 80 institutions in a coordinated manner with the goal of 
mapping the entire mutational landscape of 37 common cancer types. 
Currently each sample requires seven hours to download even on an institutional 
connection. 
Importantly, researchers do not trust the provided mapping, and thus they redo mappings. 
The time spent on mapping is about 50\% of the overall time spent on the sequence analysis pipeline. 
As read mapping is typically the most costly step in NGS analysis pipelines (e.g., GATK), any improvement to existing mappers will immediately accelerate sequence analysis studies on large read datasets.

@@

Given a sequenced genome, the next natural questions ask what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{berger2013computational} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function is
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known. 
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.
The homology search problem was previously believed to be solved;
Basic Local Alignment Search Tool (BLAST)~\cite{berger2013computational} has been
the standard tool for performing homology (similarity) search on databases of nucleotide and
protein sequences.
BLAST takes a `seed-and-extend' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running-time scales linearly with the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

@@

Techniques for reference-based read mapping typically rely on algorithmic 
approaches such as the Burrows-Wheeler transform (BWT), 
which provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure is a compressed substring index, 
based on the BWT, which provides efficient storage as well as fast search~\cite{berger2013computational}.
BWA (Burrows-Wheeler Aligner) uses the BWT, while the 
Bowtie2 mapper further relies on the FM-index for 
efficient mapping of NGS reads~\cite{berger2013computational} .
The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
coupled with dynamic programming in a compressed representation of the 
reference genome, in order to prune the search space 
when mapping reads to a reference genome.
Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2010mrsfast} use an `approximate seed' approach to index
the space of possible matches, likewise pruning the search space; however, the
bulk of its runtime is spent on the extend phase.
State-of-the-art mapper mrsFAST-Ultra achieves improvements in efficiency based on machine architecture rather than leveraging redundancy in the data itself with near-perfect sensitivity, but
only for the case where there are no insertions and deletions (indels) \cite{hach2014mrsfast}.
Even with these approaches, read mapping remains a significant bottleneck in
genomic research~\cite{berger2013computational}.

Compressing reads for storage is necessary should researchers wish to apply more advanced mapping tools or other analysis in the future.
Taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common short---15bp---substring) together into `buckets', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This approach allows a general-purpose compressor to achieve better compression.
SCALCE~\cite{berger2013computational} also relies on a `boosting' scheme, 
reordering
reads in such a way that a general-purpose compressor achieves improved
compression.

Recent advances in metagenomic search tools have relied on two improvements over
BLASTX: indexing and alphabet reduction.
RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
collision-free hash table.
The alphabet reduction, as it is reversible, can be thought of as a form of
lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
alphabet, with offsets stored to recover the original sequence in the full
alphabet.
The hash-table provides an efficient index of the database to be searched.
DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
`shaped seeds'---essentially, $k$-mers of length 15--24 with wildcards at 9--12 specific positions---instead of simple $k$-mer seeds to index the database.
DIAMOND demonstrates search performance three to four orders of magnitude faster
than BLASTX, but still linear in the size of the database being searched.

@@

\begin{table}
\label{fractal}
\caption{Metric-entropy ratio (ratio of clusters to entries in database) and
fractal dimension at typical search radii for four data sets. Metric-entropy ratio gives an estimate of the acceleration of coarse search with respect to na\"ive search, and as long as fractal dimension is low, coarse search should dominate total search time.
NCBI's non-redundant `NR' protein and `NT' nucleotide sequence databases are from June, 2015. Protein Data Bank (PDB) is from July, 2015. PubChem is from October, 2013.}
\tabcolsep=0.11cm
\begin{tabular}{H{3.0cm}p{2.2cm}p{1.8cm}}
\hline
Data set & Metric-entropy ratio & Fractal dimension \\        
\hline
Nucleotide~sequences (NCBI NT) & 7:1 & 1.5\\
\hline
Protein~sequences (NCBI NR) & 5:1 & 1.6\\
\hline
Protein~structure (PDB) & 10:1 & 2.5\\
\hline
Chemical~structure (PubChem) & 11:1 & 0.2\\
\hline
\end{tabular}
\end{table}

@@

We are entering the age of compressive algorithms, 
which make use of this completely different paradigm for the structure of biological data.
Seeking to take advantage of the redundancy inherent in genomic sequence data, 
Loh, Baym and Berger \cite{loh2012compressive}
introduced \emph{compressive genomics}, an approach that relies on compressing
data in such a way that the desired computation (such as BLAST search) can be
performed in the compressed representation.
Compressive genomics is based on the concept of \emph{compressive acceleration},
which relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
Coarse search is performed only on the coarse, or representative, subsequences
that represent unique data.
Any representative sequence within some threshold of the query is then expanded 
into all similar sequences it represents; the fine search is over this
(typically small) subset of the original database.
This approach provides orders-of-magnitude run-time improvements to BLAST 
nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
search; these run-time improvements increase as databases grow.

The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST (compressively-accelerated BLAST)~\cite{loh2012compressive} and 
caBLASTP~\cite{daniels2013compressive}, accelerates existing tools (in this 
case, read
mappers including BWA or Bowtie2) by allowing them to operate in a compressed
space, and relies on a coarse and a fine phase.
In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{berger2013computational}, conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as `seeds' for longer sequence matches.
CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each 
one to its neighbors within a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.
As an aside, though not an example of accelerated similarity search, mid-size $l$-mers
can also be used for storage compression of NGS quality scores associated with base calls.
Unlike many other approaches to compressing quality scores in the literature
~\cite{bonfield2013compression,ochoa2013qualcomp},
Quartz~\cite{yu2015quality} takes advantage of the fact that mid-size $l$-mers can in many cases almost uniquely identify locations in the genome,
bounding the likelihood that a quality score is informative and allowing for lossy compression of uninformative scores.
Because Quartz's lossy compression injects information from the distribution of $l$-mers in the target genome, it demonstrates not only improved compression over competing
approaches, but slightly improves the accuracy of downstream 
variant-calling~\cite{yu2015quality}.

@@

The compressive omics approach can, in some cases, come at the cost of accuracy.
However, these cases are well-defined.
Compressive omics never results in false positives (with respect to the na\"ive 
search technique being accelerated), because the fine search phase applies the
same comparison to the candidates as the na\"ive approach.
Furthermore, when the distance function used for comparisons is a metric---more specifically, when it obeys the triangle inequality---false negatives will also
never occur.
Yet, in practice, non-metric distance functions are used, such as E-values
in BLAST or cosine distance in esFragBag, and thus false negatives can occur.
Fortunately, these error rates are low, and recall better than 90\% has been
demonstrated~\cite{loh2012compressive,daniels2013compressive,yu2015entropy}.

@@

The approach of compressive acceleration~\cite{loh2012compressive}, and its 
demonstrated ability to scale with the metric entropy of the 
data~\cite{yu2015entropy}, while providing orthogonal benefits to many other
useful indexing techniques, is an important tool for coping with the deluge of
data.
The extension of this compressive acceleration approach to 
metagenomics, NGS read 
mapping~\cite{yorukoglu2015compressive}, and chemogenomics suggests its 
flexibility.
Likewise, compressive storage for these applications can be shown to scale with 
the information-theoretic entropy of the data set~\cite{yu2015entropy}.
